\chapter*{Acknowledgments}
\markboth{Acknowledgments}{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

I like to thank all the people with whom I had the pleasure of interacting with during my PhD at CP3, CERN, or on conferences.

A big thanks goes to Andrea Giammanco for his scientific guidance, availability, constructive feedbacks, and his open and friendly attitude. In particular I enjoyed the given freedom to peruse my interests and ideas in single-top-quark physics. I like to thank also my PhD committee members, Freya Blekman, Vincent Lemaitre, Fabio Maltoni, and Andreas Meyer for agreeing to be in my jury and for their friendly input and suggestions.

Furthermore, I like to say thanks to my colleagues, the ``The Offbeat Physicists'' team, Georgios Krintiras, Hamed Bakhshiansohi, Abideh ``Nadjieh'' Jafari, Pieter David, and Andrey Popov with whom I had a lot of fun discussing about physics and life during my PhD.

For the successful collaboration on the top quark polarization measurement I like to thank Andres Tiko and Joosep Pata for all the hard work which condensed into the published paper.

I also like to express by gratitude to the CMS (single) top conveners, Rebeca Gonzalez Suarez, Alberto Orso Maria Iorio, Abideh ``Nadjieh'' Jafari, Jeremy Andrea, Thorsten Chwalek, Andreas Meyer, Andrea Giammanco, and Pedro da Silva. It was a pleasure contributing to the single-top-quark efforts of CMS in the friendly and welcoming atmosphere you created.

I like to give my thanks also to Sezen Sekmen and Lukas Vanelderen for the great collaboration and fun we had while developing the CMS ``Fast Simulation'' package.

At CP3 I thank Carinne Mertens and Ginette Tabordon for their support when dealing with bureaucracy and travel reimbursements. I like to thank also J\'er\^ome de Favereau and Pavel Demin for their constant efforts to keep the IT infrastructure working and who always helped in fixing my grid/cluster/storage problems.

During my PhD studies I was a recipient of the FRIA scholarship by Fonds de la Recherche Scientifique, Belgium. I am grateful to the committee for awarding me the scholarship.

Finally, I like to thank my parents, sister, and friends in Germany for supporting me from far away.
